Delocalization of Eigenvectors in Random Matrices

### 2025-03-20 Draft.tex
typo

\noindent This is equivalent to $\mathbf{u}_\alpha$ being eigenvectors for the resolvent with $\frac{1}{\lambda_\alpha-z}$ as the eigenvalues. This also means that we have by spectral decomposition 
$$G(z) =\sum_{i=1}^N \frac{\mathbf{u}_\alpha\mathbf{u}_\alpha^*}{\lambda_\alpha-z}.$$
Now, recall that their empirical measure for $H$ is defined as \cite{dynamic}: $$\varrho_N(dx):=\frac{1}{N}\sum_{j=1}^N \delta(x-\lambda_j)dx,$$
Let $m(z)=m_N(z) = \int_\R \frac{\varrho_N(x)}{x-z}dx$ be its Stieltjes transform for $z = E+i\eta$, where $E\in \R$ and $\eta>0$. By the relationship we found above, we have the following  identity, which follows from the definition of the Dirac delta function: $$m(z)= \int_\R \frac{\varrho_N(x)}{x-z}dx = \int_\mathbb{R} \frac{1}{N}\frac{\sum_{j=1}^N \delta(x-\lambda_j)dx}{x-z}dx = \frac{1}{N}\sum_{j=1}^N\int_\R\frac{\delta(x-\lambda_j)}{x-z}dx =\frac{1}{N} \sum_{j=1}^N \frac{1}{\lambda_j-z} = \frac{1}{N}\text{Tr}(G(z))$$
From the moments method \cite{dynamic} we know that the Stieltjes transform $m_N(z)$ of the empirical distribution converges in probability to the Stieltjes transform $m_{SC} = \frac{1}{2}(-z+\sqrt{z^2-4})$ of the semicircle law as $N\rightarrow\infty$. This means that in the limit, with probability one the self-consistent equation holds: \begin{equation}
m(z) = -\frac{1}{z+m(z)} \Leftrightarrow m(z) + \frac{1}{m(z)} + z=0
\end{equation}
\subsection*{Useful identities}
Let us establish several facts and lemmas that we will need throughout our later work. We can begin by observing that: $$m(z)(m(z)+z) = \frac{1}{4}(-z+\sqrt{z^2-4})(z+\sqrt{z^2-4})=\frac{1}{4}(-z^2+z^2-4)=-1$$
$$\Rightarrow |m(z)| = \frac{1}{|m(z)+z|^{-1}}$$
Now, take the imaginary part of $(2)$: $$\text{Im}(m(z)) + \text{Im}\left(\frac{1}{m(z)}\right)+\eta=0$$
We can calculate the second term by simply expanding: $$\frac{1}{m(z)} = \frac{1}{\text{Re}(m(z))+\text{Im}(m(z))i} = \frac{\text{Re}(m(z))-\text{Im}(m(z))i}{|m(z)|^2}\Leftrightarrow \text{Im}\left(\frac{1}{m(z)}\right) = -\frac{\text{Im}(m(z))}{|m(z)|^2}$$
$$\Rightarrow\text{Im}(m(z)) + \text{Im}\left(\frac{1}{m(z)}\right)+\eta=\text{Im}(m(z)) -\frac{\text{Im}(m(z))}{|m(z)|^2}+\eta=0$$
$$\Rightarrow \text{Im}(m(z))\left(1-\frac{1}{|m(z)|^2}\right)=-\eta$$
And since $\text{Im}(m(z))>0$ and $\eta>0$, we must have $1-\frac{1}{|m(z)|^2}<0$, i.e $|m(z)|<1$, which is equivalent to $\text{Im }m \asymp 1$, where $\asymp$ means being bounded above and below up to a fixed, positive factor. Furthermore, let us rewrite the latter equation as the following two identities: \begin{equation*}1-|m(z)|^2 = \frac{\eta|m(z)|^2}{\text{Im}(m(z))}
\tag{3.1}
\end{equation*}
\begin{equation*}\text{Im}(m(z))= \frac{\eta|m(z)|^2}{1-|m(z)|^2 }
\tag{3.2}
\end{equation*}
Now, let $E\in [-10, 10]$ and $\eta\in (0, 10]$. First observe that $m(z)$ can never be zero, since it implies: $$m(z) = \frac{-z+\sqrt{z^2-4}}{2}=0 \Leftrightarrow \sqrt{z^2-4}=z$$
$$\Rightarrow z^2-4=z^2\Leftrightarrow -4=0$$ 
Then, clearly $\exists c>0$, s.t. $|m(z)|\geq c$. \textcolor{red}{Write lemma A.1 here or transfer the latter to the appendix}.

@@

\newpage
\section{Flow Method}
\label{sec:flow}
Consider the Brownian perturbation of the standard deviation matrix defined as $H(t)_{xy} = \sqrt{S_{xy}}b(t)_{xy}$, where $b(t)\sim \mathcal{N}(0, t)$ represents the Brownian motion. For this system, the matrix resolvent is augmented as $G(t) = (H(t)-w(t))^{-1}$ , where $w(t) = -\frac{1}{m(z)}-tm(z)$. In order to study this system and its transition from $t=0$ with $G(0) = m(z)$, to $t=1$ with $G(1)=G(z)$, we will employ stochastic calculus. For this purpose, observe that our differentials are $dH(t)_{xy} = \sqrt{S_{xy}}db_{xy}(t)$ and $\partial_tw(t) = -m(z)$. Recall the Îto equation, namely, the statement that a function $f(\vec{b}(t), t)$, dependent on a Brownian random walk  $\vec{b}(t) = (b(t)_1, \cdots, b(t)_d)$ w.r.t to time $t$, has derivative:
$$df(\vec{b}(t), t) = \frac{\partial f}{\partial t}dt + \sum_{i=1}^d \frac{\partial f}{\partial b_i}db_i(t) + \frac{1}{2}\sum_{i=1}^d \frac{\partial^2 f}{\partial b_i^2} dt$$
In our case, by taking the function to be $G(t)$ and using restriction of a symmetric matrix and the latter differentials, we get the following expression:

@@

\subsection{Dynamical $T$-equation}
Now, define $F_{xy} = |G(t)_{xy}|^2 = G(t)_{xy}\overline{G(t)_{xy}}$. 
 Applying Itô's formula to \( F(t)_{xy} \), we have:
\[
dF(t)_{xy}= d|G_{xy}|^2= G(t)_{xy} d\overline{G(t)}_{xy} + \overline{G(t)}_{xy} dG(t)_{xy} + dG(t)_{xy} d\overline{G(t)}_{xy}.
\]
$$d|G_{xy}|^2 = G_{xy}d\overline{G_{xy}} + \overline{G}_{xy}dG_{xy} + d\left[G_{xy}, \overline{G}_{xy}\right]$$
 $$d|G_{xy}|^2 = G_{xy}d\overline{G_{xy}} + \overline{G}_{xy}dG_{xy} +d\left[\sum_{\alpha, \beta} G_{x\alpha}d H_{\alpha \beta}\overline{G}_{xy}\right]$$
$$d|G_{xy}|^2 = G_{xy}d\overline{G_{xy}} + \overline{G}_{xy}dG_{xy} +\sum_{\alpha,\beta, \gamma, \delta}G_{x\alpha}G_{\beta\gamma}\overline{G_{xy}}\overline{G_{\delta \gamma}}$$
To get the $T$-equation, define it as:
\[
T(t) =  S^{1/2}F(t)S^{1/2}=\sum_{x,y} S_{xy} F(t)_{xy} = \sum_{x,y} S_{xy} |G(t)_{xy}|^2,
\]
In order to find its derivative, we need to substitute for the earlier terms we found, namely \( dG(t)_{xy} = dI(t)_{xy} + dII(t)_{xy} dt \) and its conjugate, noting that \( dI(t)_{xy} d\overline{I(t)}_{xy} \) contributes to the quadratic variation, we obtain
\(
dF(t)_{xy} = dM(t)_{xy} + \Omega(t)_{xy} dt
\), where
\begin{equation*}
dM(t)_{xy} = \overline{G(t)}_{xy} dI(t)_{xy} + G(t)_{xy} \overline{dI(t)}_{xy} \end{equation*}
\begin{equation*}
\Omega(t)_{xy} = G_{xy}\frac{\overline{dII(t)}_{xy}}{dt}+\overline{G_{xy}}\frac{dII(t)_{xy}}{dt}+\frac{d\left[G_{xy}, \overline{G_{xy}}\right]}{dt}
\end{equation*}
and when expanded w.r.t to $dI(t)$ and $dII(t)dt$:
$$dM(t)_{xy} = \sum_{\alpha, \beta}\overline{G_{xy}}G_{x\alpha}{S_{\alpha\beta}}^{1/2}G_{\beta y}dB_{\beta(t)}+\sum_{\alpha, \beta}{G}_{xy}\overline{G_{x\alpha}}{S_{\alpha\beta}}^{1/2}\,\overline{G_{\beta y}}dB_{\beta(t)}$$

$$\Omega(t)_{xy}=G_{xy}(t)\left(\overline{\sum_j G_{xj}(t)S_{jy}(G_{yy}(t)-m(z))G_{jy}(t)}\right)+$$
$$+\overline{G_{xy}(t)}\left(\sum_j G_{xj}(t)S_{jy}(G_{yy}(t)-m(z))G_{jy}(t)\right)+\sum_{\alpha, \beta}|G_{xa}(t)|^2S_{\alpha, \beta}|G_{x\beta}(t)|^2$$
Where by Itô, the $T$-equation becomes: 
\begin{equation}
\Rightarrow d{T} = {T}^2 dt + S^{1/2} dM_t(z)S^{1/2} + S^{1/2} \Omega_t(z) S^{1/2}dt
\end{equation}
Now, define the \textit{time-dependent diffusion profile} as:$$\Theta_t = \frac{|m(z)|^2S}{1-t|m(z)|^2S}$$
Let us take its derivative w.r.t to $t$: 
$$\frac{d\Theta_t}{dt} = \frac{d}{dt}\frac{|m(z)|^2S}{1-t|m(z)|^2S} = \frac{-|m(z)|^2S(-|m(z)|^2S)}{(1-t|m(z)|^2S)^2} = \Theta_t^2 \Leftrightarrow d\Theta_t = \Theta_t^2 dt$$

@@

\noindent The reasoning behind the latter definitions is simple - informally, our goal is to establish a bootstrapping mechanism that will allow us to propagate forward the stopping time through a self-reinforcement (continuity) argument. While the particular choices for constants may seem arbitrary at the moment, they are informed by the bounds we get on the maximal entries of the diffusion profile $\Theta_t$ (\hyperref[diffusion]{Sec. 4.2}), the assumption of $\eta\gg W^{-3/4}$, as well as the technical steps in our analytic argument. That said, we can still gain some intuition as to the meaning of each. With $\tau_{\text{stop},1}$ we have control over the size the error term $E_s(z) = T_s(z) - \Theta_s$, whereas $\tau_{\text{stop},2}$ is w.r.t to the deviation of the resolvent entries from the semicircle law $m(z)$, normalized by appropriate terms. The self-reinforcing argument works as follows:

@@

\noindent The particular delocalization bound for $W \gg N^{8/11}$ emerges from the need of ensuring that the combined error term $W^{-\frac{7}{4}}\eta^{-\frac{3}{2}}$ in equation $(3.2.1)$ is much smaller than the typical size of the $\Theta$ entries of $O( W^{-1}\eta^{-\frac{1}{2}})$, while maintaining he diffusion time scale $\eta \approx W^2N^{-2}$ (\hyperref[diffusion]{Sec. 4.2}) :

@@

@@

\subsection{Quantum Diffusion}
\label{diffusion}

Hence, our $\Theta$ term simplify $$\Theta \sim \frac{S}{1-|m(z)|^2+|m(z)|^2(S-\text{Id})}\sim \frac{S}{\alpha\eta+(S-\text{Id})}$$

>>>>

\noindent This is equivalent to $\mathbf{u}_\alpha$ being eigenvectors for the resolvent with $\frac{1}{\lambda_\alpha-z}$ as the eigenvalues. This also means that we have by spectral decomposition 
$$G(z) =\sum_{i=1}^N \frac{\mathbf{u}_\alpha\mathbf{u}_\alpha^*}{\lambda_\alpha-z}.$$
Now, recall that their empirical measure for $H$ is defined as \cite{dynamic}: $$\varrho_N(dx):=\frac{1}{N}\sum_{j=1}^N \delta(x-\lambda_j)dx,$$
Let $m(z)=m_N(z) = \int_\R \frac{\varrho_N(x)}{x-z}dx$ be its Stieltjes transform for $z = E+i\eta$, where $E\in \R$ and $\eta>0$. By the relationship we found above, we have the following  identity, which follows from the definition of the Dirac delta function: $$m(z)= \int_\R \frac{\varrho_N(x)}{x-z}dx = \int_\mathbb{R} \frac{1}{N}\frac{\sum_{j=1}^N \delta(x-\lambda_j)dx}{x-z}dx = \frac{1}{N}\sum_{j=1}^N\int_\R\frac{\delta(x-\lambda_j)}{x-z}dx =\frac{1}{N} \sum_{j=1}^N \frac{1}{\lambda_j-z} = \frac{1}{N}\text{Tr}(G(z))$$
From the moments method \cite{dynamic} we know that the Stieltjes transform $m_N(z)$ of the empirical distribution converges in probability to the Stieltjes transform $m_{SC} = \frac{1}{2}(-z+\sqrt{z^2-4})$ of the semicircle law as $N\rightarrow\infty$. This means that in the limit, with probability one the self-consistent equation holds: 
\label{self-consistent}
\begin{equation}
m(z) = -\frac{1}{z+m(z)} \Leftrightarrow m(z) + \frac{1}{m(z)} + z=0
\end{equation}

\subsection*{Useful identities}
Let us establish several facts and lemmas that we will need throughout our later work. We can begin by observing that: $$m(z)(m(z)+z) = \frac{1}{4}(-z+\sqrt{z^2-4})(z+\sqrt{z^2-4})=\frac{1}{4}(-z^2+z^2-4)=-1$$
$$\Rightarrow |m(z)| = \frac{1}{|m(z)+z|^{-1}}$$
Now, take the imaginary part of $(2)$: $$\text{Im}(m(z)) + \text{Im}\left(\frac{1}{m(z)}\right)+\eta=0$$
We can calculate the second term by simply expanding: $$\frac{1}{m(z)} = \frac{1}{\text{Re}(m(z))+\text{Im}(m(z))i} = \frac{\text{Re}(m(z))-\text{Im}(m(z))i}{|m(z)|^2}\Leftrightarrow \text{Im}\left(\frac{1}{m(z)}\right) = -\frac{\text{Im}(m(z))}{|m(z)|^2}$$
$$\Rightarrow\text{Im}(m(z)) + \text{Im}\left(\frac{1}{m(z)}\right)+\eta=\text{Im}(m(z)) -\frac{\text{Im}(m(z))}{|m(z)|^2}+\eta=0$$
$$\Rightarrow \text{Im}(m(z))\left(1-\frac{1}{|m(z)|^2}\right)=-\eta$$
And since $\text{Im}(m(z))>0$ and $\eta>0$, we must have $1-\frac{1}{|m(z)|^2}<0$, i.e $|m(z)|<1$, which is equivalent to $\text{Im }m \asymp 1$, where $\asymp$ means being bounded above and below up to a fixed, positive factor. Furthermore, let us rewrite the latter equation as the following two identities: \begin{equation*}1-|m(z)|^2 = \frac{\eta|m(z)|^2}{\text{Im}(m(z))}
\tag{3.1}
\end{equation*}
\begin{equation*}\text{Im}(m(z))= \frac{\eta|m(z)|^2}{1-|m(z)|^2 }
\tag{3.2}
\end{equation*}
Now, let $E\in [-10, 10]$ and $\eta\in (0, 10]$. First observe that $m(z)$ can never be zero, since it implies: $$m(z) = \frac{-z+\sqrt{z^2-4}}{2}=0 \Leftrightarrow \sqrt{z^2-4}=z$$
$$\Rightarrow z^2-4=z^2\Leftrightarrow -4=0$$ 
Then, clearly $\exists c>0$, s.t. $|m(z)|\geq c$. \textcolor{red}{Write lemma A.1 here or transfer the latter to the appendix}.

@@

\newpage
\section{Flow Method}
\label{sec:flow}
Consider the Brownian perturbation of the standard deviation matrix defined as $H(t)_{xy} = \sqrt{S_{xy}}b(t)_{xy}$, where $b(t)\sim \mathcal{N}(0, t)$ represents the Brownian motion. For this system, the matrix resolvent is augmented as $G(t) = (H(t)-w(t))^{-1}$ , where: $$w(t) = -\frac{1}{m(z)}-tm(z) = z+(1-t)m(z)$$
by the self-consistent equation \hyperref[self-consistent]{(2)}. In order to study this system and its transition from $t=0$ with $G(0) = m(z)$, to $t=1$ with $G(1)=G(z)$, we will employ stochastic calculus. For this purpose, observe that our differentials are $dH(t)_{xy} = \sqrt{S_{xy}}db_{xy}(t)$ and $\partial_tw(t) = -m(z)$. Recall the Îto equation, namely, the statement that a function $f(\vec{b}(t), t)$, dependent on a Brownian random walk  $\vec{b}(t) = (b(t)_1, \cdots, b(t)_d)$ w.r.t to time $t$, has derivative:
$$df(\vec{b}(t), t) = \frac{\partial f}{\partial t}dt + \sum_{i=1}^d \frac{\partial f}{\partial b_i}db_i(t) + \frac{1}{2}\sum_{i=1}^d \frac{\partial^2 f}{\partial b_i^2} dt$$
In our case, by taking the function to be $G(t)$ and using restriction of a symmetric matrix and the latter differentials, we get the following expression:

@@

\subsection{Dynamical $T$-equation}
Having found the differential of the resolvent, we will now proceed with defining the time-dependent $T$-matrix as:
$$T(t)_{ab} = \sum_{x,y}S_{ax}^{\frac{1}{2}}F(t)_{xy}S_{yb}^{\frac{1}{2}},$$
where $F(t)_{xy} = |G(t)_{xy}|^2 = G(t)_{xy}\overline{G(t)_{xy}}$. 
\textcolor{red}{Applying Itô's formula}(show it explicitly) to \( F(t)_{xy} \), we have:
\[
dF(t)_{xy}= d|G_{xy}|^2= G(t)_{xy} d\overline{G(t)}_{xy} + \overline{G(t)}_{xy} dG(t)_{xy} + dG(t)_{xy} d\overline{G(t)}_{xy}.
\]
$$d|G_{xy}|^2 = G_{xy}d\overline{G_{xy}} + \overline{G}_{xy}dG_{xy} + d\left[G_{xy}, \overline{G}_{xy}\right]$$
 $$d|G_{xy}|^2 = G_{xy}d\overline{G_{xy}} + \overline{G}_{xy}dG_{xy} +d\left[\sum_{\alpha, \beta} G_{x\alpha}d H_{\alpha \beta}\overline{G}_{xy}\right]$$
$$d|G_{xy}|^2 = G_{xy}d\overline{G_{xy}} + \overline{G}_{xy}dG_{xy} +\sum_{\alpha,\beta, \gamma, \delta}G_{x\alpha}G_{\beta\gamma}\overline{G_{xy}}\overline{G_{\delta \gamma}}$$
In order to find the derivative of the $T$-matrix, we need to substitute for the earlier terms we found, namely \( dG(t)_{xy} = dI(t)_{xy} + dII(t)_{xy} dt \) and its conjugate, noting that \( dI(t)_{xy} d\overline{I(t)}_{xy} \) contributes to the quadratic variation, we obtain
\(
dF(t)_{xy} = dM(t)_{xy} + \Omega(t)_{xy} dt
\), where
\begin{equation*}
dM(t)_{xy} = \overline{G(t)}_{xy} dI(t)_{xy} + G(t)_{xy} \overline{dI(t)}_{xy} \end{equation*}
\begin{equation*}
\Omega(t)_{xy} = G_{xy}\frac{\overline{dII(t)}_{xy}}{dt}+\overline{G_{xy}}\frac{dII(t)_{xy}}{dt}+\frac{d\left[G_{xy}, \overline{G_{xy}}\right]}{dt}
\end{equation*}
and when expanded w.r.t to $dI(t)$ and $dII(t)dt$:
$$dM(t)_{xy} = \sum_{\alpha, \beta}\overline{G_{xy}}G_{x\alpha}{S_{\alpha\beta}}^{1/2}G_{\beta y}dB_{\beta(t)}+\sum_{\alpha, \beta}{G}_{xy}\overline{G_{x\alpha}}{S_{\alpha\beta}}^{1/2}\,\overline{G_{\beta y}}dB_{\beta(t)}$$

$$\Omega(t)_{xy}=G_{xy}(t)\left(\overline{\sum_j G_{xj}(t)S_{jy}(G_{yy}(t)-m(z))G_{jy}(t)}\right)+$$
$$+\overline{G_{xy}(t)}\left(\sum_j G_{xj}(t)S_{jy}(G_{yy}(t)-m(z))G_{jy}(t)\right)+\sum_{\alpha, \beta}|G_{xa}(t)|^2S_{\alpha, \beta}|G_{x\beta}(t)|^2$$
Hence, by Itô the $T$-equation becomes: 
\begin{equation}
\Rightarrow d{T} = {T}^2 dt + S^{1/2} dM_t(z)S^{1/2} + S^{1/2} \Omega_t(z) S^{1/2}dt
\end{equation}
Now, define the \textit{time-dependent diffusion profile} as:$$\Theta_t = \frac{|m(z)|^2S}{1-t|m(z)|^2S}$$
Let us take its derivative w.r.t to $t$: 
$$\frac{d\Theta_t}{dt} = \frac{d}{dt}\frac{|m(z)|^2S}{1-t|m(z)|^2S} = \frac{-|m(z)|^2S(-|m(z)|^2S)}{(1-t|m(z)|^2S)^2} = \Theta_t^2 \Leftrightarrow d\Theta_t = \Theta_t^2 dt$$

@@

\noindent The reasoning behind the latter definitions is simple - informally, our goal is to establish a bootstrapping mechanism that will allow us to propagate forward the stopping time through a self-reinforcement (continuity) argument. While the particular choices for constants may seem arbitrary at the moment, they are informed by the bounds we get on the maximal entries of the diffusion profile $\Theta_t$ (\hyperref[diffusion]{Sec. 4.1}), the assumption of $\eta\gg W^{-3/4}$, as well as the technical steps in our analytic argument. That said, we can still gain some intuition as to the meaning of each. With $\tau_{\text{stop},1}$ we have control over the size the error term $E_s(z) = T_s(z) - \Theta_s$, whereas $\tau_{\text{stop},2}$ is w.r.t to the deviation of the resolvent entries from the semicircle law $m(z)$, normalized by appropriate terms. The self-reinforcing argument works as follows:

@@

\noindent The particular delocalization bound for $W \gg N^{8/11}$ emerges from the need of ensuring that the combined error term $W^{-\frac{7}{4}}\eta^{-\frac{3}{2}}$ in equation $(3.2.1)$ is much smaller than the typical size of the $\Theta$ entries of $O( W^{-1}\eta^{-\frac{1}{2}})$, while maintaining he diffusion time scale $\eta \approx W^2N^{-2}$ (\hyperref[diffusion]{Sec. 4.1}) :

@@

\subsection{Quantum Diffusion}
\label{diffusion}
Recall our time-dependent $T$-equation $T(t)$ and diffusion profile $\Theta_t$. The $T$-matrix and diffusion profile $\Theta$ are just the latter two for $t=1$, i.e: $$T(z)_{ab} = T_{ab} = \sum_{x,y}S_{ax}^{\frac{1}{2}} |G_{xy}|^2S_{yb}^{\frac{1}{2}}$$
Hence, our $\Theta$ term simplify $$\Theta \sim \frac{S}{1-|m(z)|^2+|m(z)|^2(S-\text{Id})}\sim \frac{S}{\alpha\eta+(S-\text{Id})}$$

@@
